### 2015-07-31 implementation-environment.tex
Changes from Greg.

Argument parsing and indirection cause Spack's compiler wrappers to incur a small
but noticeable performance overhead.  Figure~\ref{fig:build-time} shows
build times with and without compiler wrappers for seven different packages,
and Figure~\ref{fig:overhead} shows the corresponding overhead as a percentage
of the wrapper-less runtime.  {\tt Dyninst} and {\tt lapack}
use CMake to build, and the rest use autotools.  We ran
each build three times on an Intel Sandy Bridge cluster node, and we
report the average here.
In the case of {\tt dyninst}, the overhead is negligible (-0.4\%),
but it is as high as 12.3\% for shorter builds, like
{\tt mpileaks}.

To increase build speed, by default Spack runs each {\tt install}
method in the default temporary directory. On most cluster nodes,
this is a fast, locally-mounted filesystem. In our experience,
many users build manually in their home directories out of habit.  At most sites, home
directories are remotely mounted volumes (e.g., NFS).  To compare Spack builds
with {\it typical} manual builds, we timed the same seven builds using
NFS.  Figure~\ref{fig:overhead} shows that building this way can be as much
as 62.7\% slower than using a temporary filesystem and
33\% slower on average. We therefore expect many users to notice
a speedup when Spack uses temporary space to build.

\subsubsection{Environment Module Integration}
\label{sec:envmodule}
In addition to managing the build-time environment, Spack can assist in managing
the run-time environment.  A package may need environment variables like {\tt PATH},
{\tt MANPATH}, or {\tt PKG\_CONFIG\_PATH} set before it can be used.
As discussed in Section~\ref{sec:motivation}, many sites rely on environment
modules to setup the runtime environment.  Spack can automatically create simple
dotkit~\cite{dotkit} and Module configuration files for its packages, allowing
users to setup their runtime environment using familiar systems.  While Spack
packages do not {\tt LD\_LIBRARY\_PATH} to run, we set it in our generated module
files as well, because it can be used by build systems to find libraries, as well
as by dependent packages that do not use {\tt RPATH}.
%
Future versions of Spack may also allow the creation of Lmod~\cite{mclay:lmod}
hierarchies, as discussed in Section~\ref{sec:env-rpath}. Spack's rich
dependency information would allow such hierarchies to be generated automatically.

>>>>

Argument parsing and indirection cause Spack's compiler wrappers to incur a small
but noticeable performance overhead.  Figure~\ref{fig:build-times} shows
build times with and without compiler wrappers for seven different packages,
and Figure~\ref{fig:overhead} shows the corresponding overhead as a percentage
of the wrapper-less runtime.  {\tt Dyninst} and {\tt LAPACK}
use CMake to build, and the rest use autotools.  We ran
each build three times on an Intel Sandy Bridge cluster node, and we
report the average here.
In the case of {\tt dyninst}, the overhead is negligible (-0.4\%),
but it is as high as 12.3\% for shorter builds, like
{\tt mpileaks}.

To increase build speed, by default Spack runs each {\tt install}
method in the default temporary directory. On most cluster nodes,
this is a fast, locally-mounted file system. In our experience,
many users build manually in their home directories out of habit.  At most sites, home
directories are remotely mounted volumes (e.g., NFS).  To compare Spack builds
with {\it typical} manual builds, we timed the same seven builds using
NFS.  Figure~\ref{fig:overhead} shows that building this way can be as much
as 62.7\% slower than using a temporary file system and
33\% slower on average. We therefore expect many users to notice
a speedup when Spack uses temporary space to build.

\subsubsection{Environment Module Integration}
\label{sec:envmodule}
In addition to managing the build-time environment, Spack can assist in managing
the run-time environment.  A package may need environment variables like {\tt PATH},
{\tt MANPATH}, or {\tt PKG\_CONFIG\_PATH} set before it can be used.
As discussed in Section~\ref{sec:motivation}, many sites rely on environment
modules to setup the runtime environment.  Spack can automatically create simple
dotkit~\cite{dotkit} and Module configuration files for its packages, allowing
users to setup their runtime environment using familiar systems.  While Spack
packages do not need {\tt LD\_LIBRARY\_PATH} to run, we set it in our generated module
files as well, because it can be used by build systems to find libraries, as well
as by dependent packages that do not use {\tt RPATH}.
%
Future versions of Spack may also allow the creation of Lmod~\cite{mclay:lmod}
hierarchies, as discussed in Section~\ref{sec:env-rpath}. Spack's rich
dependency information would allow such hierarchies to be generated automatically.

### 2015-07-31 implementation-packages.tex
Changes from Greg.

In Spack, packages are Python scripts that build software artifacts.
Each package is a class that extends a generic {\tt Package}
base class.  {\tt Package} implements the bulk of the build process, but
subclasses provide their own {\tt install} method to handle the
specifics of particular packages. The subclass does {\it not} have to
handle managing the install location.  Rather, Spack passes the {\tt install}
method a {\tt prefix} parameter.  The package implementor must ensure that
her {\tt install} function installs the package {\it into} the {\tt prefix},
but Spack ensures that {\tt prefix} is computed in such a way that it is
unique for every configuration of a package.  To further simplify
package implementation, Spack implements an embedded domain-specific
language (DSL).
The DSL provides special directives such as {\tt depends\_on},
{\tt version}, {\tt patch}, and {\tt provides} that add metadata
to the package class.

>>>>

In Spack, packages are Python scripts that build software artifacts.
Each package is a class that extends a generic {\tt Package}
base class.  {\tt Package} implements the bulk of the build process, but
subclasses provide their own {\tt install} method to handle the
specifics of particular packages. The subclass does {\it not} have to
handle managing the install location.  Rather, Spack passes the {\tt install}
method a {\tt prefix} parameter.  The package implementer must ensure that
her {\tt install} function installs the package {\it into} the {\tt prefix},
but Spack ensures that {\tt prefix} is computed so that it is
unique for every configuration of a package.  To further simplify
package implementation, Spack implements an embedded domain-specific
language (DSL).
The DSL provides directives, such as {\tt depends\_on},
{\tt version}, {\tt patch}, and {\tt provides}, that add metadata
to the package class.

### 2015-07-31 implementation-specs.tex
Changes from Greg.

{\bf Variants.}
To handle options like compiler flags or optional components, specs can
have named flags, or {\it variants}.  Variants are associated with the package,
so the {\tt mpileaks} package implementor must explicitly handle cases
where debug is enabled ({\tt +debug}) or disabled ({\tt -debug}
or {\tt \textasciitilde \ignorespaces debug}).  Names simplify versioning
and prevent the configuration space from becoming too large.
For example, including detailed compiler flags in spec syntax
would violate our goal of conciseness, but known sets of flags can simply be named.

>>>>

{\bf Variants.}
To handle options like compiler flags or optional components, specs can
have named flags, or {\it variants}.  Variants are associated with the package,
so the {\tt mpileaks} package implementer must explicitly handle cases
where debug is enabled ({\tt +debug}) or disabled ({\tt -debug}
or {\tt \textasciitilde \ignorespaces debug}).  Names simplify versioning
and prevent the configuration space from becoming too large.
For example, including detailed compiler flags in spec syntax
would violate our goal of conciseness, but known sets of flags can simply be named.

### 2015-07-31 implementation-virtual-deps.tex
Name change spack-paper-sc15 -> spack-sc15
Bibliography formatting, tighten to 12 pages.
Changes from Greg.

@@

To allow rapid composition of libraries using an interface, Spack supports
{\it virtual dependencies}.  A virtual dependency is an abstract name
representing a library interface or capability instead of a library 
implementation.  Packages that need this capability need not depend on 
a specific implementation; they can depend on the virtual name, for which
the user or Spack can select an an implementation at build time.
Other package managers support the notion of virtual dependencies, but Spack
adds {\it versioning} to its interfaces.  This allows concepts like MPI
interface versions and BLAS levels to be represented directly.  Spack 
handles the details of managing and checking complex versioning constraints.

>>>>

\begin{figure*}[ht!]
	\centering
	\includegraphics[width=.9\textwidth]{figs/concretization.pdf}
	\caption{
		Spack's concretization process.
		\label{fig:concretization}
	}
\end{figure*}

@@

To allow rapid composition of libraries using an interface, Spack supports
{\it virtual dependencies}.  A virtual dependency is an abstract name
representing a library interface or capability instead of a library
implementation.  Packages that need this capability need not depend on
a specific implementation; they can depend on the virtual name, for which
the user or Spack can select an implementation at build time.
Other package managers support the notion of virtual dependencies, but Spack
adds {\it versioning} to its interfaces.  This allows concepts like MPI
interface versions and BLAS levels to be represented directly.  Spack
handles the details of managing and checking complex versioning constraints.

### 2015-07-31 introduction.tex
Changes from Greg.

The Livermore Computing (LC) facility at Lawrence Livermore National Laboratory
(LLNL) supports around 2,500 users on 25 different clusters, ranging
in size from a 1.6 teraflop, 256-core cluster to the
20 petaflop, 1.6 million-core Sequoia machine, currently ranked first and
third on the Graph500~\cite{graph500} and Top500~\cite{top500}
lists, respectively.
%
%The massively parallel simulations that run on LC machines support
%U.S. Department of Energy missions across a range of scientific domains,
%including nuclear physics, material science, climate science, mechanical
%engineering, geophysics, seismology, and stockpile stewardship.
%
The simulation software that runs on these machines is very complex; some
codes depend on specific versions of over 70 dependency libraries.
They require specific compilers, build options and Message Passing Interface (MPI)
implementations to
achieve the best performance, and users may run several
different codes in the same environment as part of larger
scientific workflows.

>>>>

The Livermore Computing (LC) facility at Lawrence Livermore National Laboratory
(LLNL) supports around 2,500 users on 25 different clusters, ranging
in size from a 1.6 teraflop, 256-core cluster to the
20 petaflop, 1.6 million-core Sequoia machine, currently ranked second and
third on the Graph500~\cite{graph500} and Top500~\cite{top500}
lists, respectively.
%
%The massively parallel simulations that run on LC machines support
%U.S. Department of Energy missions across a range of scientific domains,
%including nuclear physics, material science, climate science, mechanical
%engineering, geophysics, seismology, and stockpile stewardship.
%
The simulation software that runs on these machines is very complex; some
codes depend on specific versions of over 70 dependency libraries.
They require specific compilers, build options and Message Passing Interface (MPI)
implementations to
achieve the best performance, and users may run several
different codes in the same environment as part of larger
scientific workflows.

### 2015-07-31 usecase-policy.tex
Name change spack-paper-sc15 -> spack-sc15
Revisions and better overhead plot.
Changes from Greg.

While Spack can easily create many installations of a package like {\tt mpileaks}
for different compilers and MPI implementations, end users may find Spack's directory layout confusing, and they may not be able to find libraries intuitively.
Spack installs packages into paths based on concretized specs.
This is ideal for maintaining multiple package installations, 
but it may be difficult for an end-user to navigate.
For example, an {\tt mpileaks} installation prefix might be:
%
\begin{minted}[fontsize=\small]{bash}
spack/opt/linux-x86_64/gcc-4.9.2/mpileaks-1.0-db465029
\end{minted}

As discussed in Section~\ref{sec:envmodule},
environment modules are commonly used on HPC systems to solve this problem,
and Spack allows the package author to automatically create environment modules
for packages that Spack installs.
Even when modules are available, many users still
navigate the filesystem to access packages.

@@

Views are configured using configuration files, which can be set up at the site or
user level.
For each package or set of packages, configuration file contains rules
describing the links that should point into that package.
The link names can be parameterized.
For example, the above {\tt mpileaks} symbolic link might have been created by a rule like:
%
\begin{minted}[fontsize=\small]{bash}
    /opt/${PACKAGE}-${VERSION}-${MPINAME}
\end{minted}
%
On installation and removal,
links are automatically created, deleted, or updated according to these rules.

Spack's views are a projection from a point in a high-dimensional space
(the concretized spec, which fully specifies all parameters) to point
in a lower-dimensional space
(the link name, which may only contain a few parameters).
Thus, several installations may map to the same link.
For example, the above {\tt mpileaks} link could point to an {\tt mpileaks} compiled with
{\tt gcc} or {\tt icc} -- the compiler parameter is not part of the link.
To keep  package installations consistent and reproducible,
Spack has a well-defined mechanism for resolving conflicting links;
it uses a combination of internal default policies and user- or site-defined
policies to define an {\it order} of preference for different parameters.
By default, Spack prefers newer versions of packages compiled with newer compilers
to older packages built with older compilers. It has well-defined, but not 
necessarily meaningful, order of preference for deciding between MPI
implementations and different compilers.
The default policies can be overridden in configuration files, by either users
or by sites. For example, at one site users may typically use the Intel compiler,
but some users also use the system's default {\tt gcc@4.4.7}.
These preferences could be stated by adding:
%
\begin{minted}[fontsize=\small]{bash}
    compiler_order = icc,gcc@4.4.7
\end{minted}
%
to the site's {\tt spackconfig} file, which would cause the ambiguous
{\tt mpileaks} link to point to an installation compiled with {\tt icc}.
Any compiler not in the compiler\_order setting is treated as less preferred
than those explicitly provided.
%
In a similar manner, Spack can be configured to give specific package
configurations priority over others.
This can be useful if a new version is unstable and untested.

@@

To support our own private packages, and to support those of LLNL code teams,
Spack allows the creation of site-specific variants of packages.
Via configuration files, users can specify additional search
directories for finding additional {\tt Package} classes.
%
The additional packages are like the {\tt mpileaks} package shown in Figure~\ref{fig:mpileaks}.  However, the extension packages can extend from not
only {\tt Package}, but also any of Spack's built-in packages.
Custom packages can inherit from and replace Spack's default
packages, so other sites to either tweak or completely replace Spack's
build recipes.  To continue the previous example, a site can write
a {\tt LocalSpindle} Python class, which inherits from Spack's
{\tt Spindle} class.  {\tt LocalSpindle} may simply add additional
configure flags to the {\tt Spindle} class, while leaving the
dependencies and most of the build instructions from its parent
class.  For reproducibility, Spack also tracks the {\tt Package}
class that drove a specific build.

>>>>

While Spack can easily create many installations of a package like {\tt mpileaks}
for different compilers and MPI implementations, end-users may find Spack's directory layout confusing, and they may not be able to find libraries intuitively.
Spack installs packages into paths based on concretized specs.
This is ideal for maintaining multiple package installations,
but it may be difficult for an end-user to navigate.
For example, an {\tt mpileaks} installation prefix might be:
%
\begin{minted}[fontsize=\small]{bash}
spack/opt/linux-x86_64/gcc-4.9.2/mpileaks-1.0-db465029
\end{minted}

As discussed in Section~\ref{sec:envmodule},
environment modules are commonly used on HPC systems to solve this problem,
and Spack allows the package author to automatically create environment modules
for packages that Spack installs.
Even when modules are available, many users still
navigate the file system to access packages.

@@

Views are configured using configuration files, which can be set up at the site or
user level.
For each package or set of packages, the configuration file contains rules
describing the links that should point into that package.
The link names can be parameterized.
For example, the above {\tt mpileaks} symbolic link might have been created by a rule like:
%
\begin{minted}[fontsize=\small]{bash}
    /opt/${PACKAGE}-${VERSION}-${MPINAME}
\end{minted}
%
On installation and removal,
links are automatically created, deleted, or updated according to these rules.

Spack's views are a projection from a point in a high-dimensional space
(the concretized spec, which fully specifies all parameters) to point
in a lower-dimensional space
(the link name, which may only contain a few parameters).
Thus, several installations may map to the same link.
For example, the above {\tt mpileaks} link could point to an {\tt mpileaks} compiled with
{\tt gcc} or {\tt icc} -- the compiler parameter is not part of the link.
To keep  package installations consistent and reproducible,
Spack has a well-defined mechanism for resolving conflicting links;
it uses a combination of internal default policies and user- or site-defined
policies to define an {\it order} of preference for different parameters.
By default, Spack prefers newer versions of packages compiled with newer compilers
to older packages built with older compilers. It has well-defined, but not
necessarily meaningful, order of preference for deciding between MPI
implementations and different compilers.
The default policies can be overridden in configuration files, by either users
or by sites. For example, at one site users may typically use the Intel compiler,
but some users also use the system's default {\tt gcc@4.4.7}.
These preferences could be stated by adding:
%
\begin{minted}[fontsize=\small]{bash}
    compiler_order = icc,gcc@4.4.7
\end{minted}
%
to the site's configuration file, which would cause the ambiguous
{\tt mpileaks} link to point to an installation compiled with {\tt icc}.
Any compiler not in the compiler\_order setting is treated as less preferred
than those explicitly provided.
%
In a similar manner, Spack can be configured to give specific package
configurations priority over others.
This can be useful if a new version is unstable and untested.

@@

To support our own private packages, and to support those of LLNL code teams,
Spack allows the creation of site-specific variants of packages.
Via configuration files, users can specify additional search
directories for finding additional {\tt Package} classes.
%
The additional packages are like the {\tt mpileaks} package shown in Figure~\ref{fig:mpileaks}.  However, the extension packages can extend from not
only {\tt Package}, but also any of Spack's built-in packages.
Custom packages can inherit from and replace Spack's default
packages, so other sites can either tweak or completely replace Spack's
build recipes.  To continue the previous example, a site can write
a {\tt LocalSpindle} Python class, which inherits from Spack's
{\tt Spindle} class.  {\tt LocalSpindle} may simply add additional
configure flags to the {\tt Spindle} class, while leaving the
dependencies and most of the build instructions from its parent
class.  For reproducibility, Spack also tracks the {\tt Package}
class that drove a specific build.
